\chapter*{Summary}
Dialysis is a treatment allowing to purify the blood from uremic toxins. Dialysis also induce a complex dynamic of fluid and substances exchange between the various compartments of the human body: from the plasma compartment, which is the most superficial, to the intracellular one, deeper and inaccessible. It is important for therapeutic purposes, to be able to describe and predict these dynamics, so that we can better tailor the therapy on every patient, by choosing among the technical options available nowadays. In fact, with respect to the recent past when the main treatment was the \textit{hemodialysis}, now is possible to choose several alternatives, among which we can find the \textit{hemodiafiltration} (HDF).
The use of mathematical modeling can help in the choice of the most suitable dialysis, and this thesis aims to provide such a tool for the special case of \textit{on-line} HDF.

The model here developed is based on the description of the kinetics of some blood solute considered to be important to monitor (sodium, potassium, chloride, calcium, inorganic phosphate, magnesium, urea, creatinine, proteins), the exchange of fluids between the different compartments, and pressure changes during a dialysis session performed with the HDF technique. In particular, we have tried to mathematically synthesized in a single system of ordinary differential equations, the interactions that occur between the following three systems:
\begin{enumerate}
	\item the ``patient'' system , described as three-compartmental (plasma, interstitial and intracellular) separated by semipermeable membranes;
	\item the ``on-line'' system for the production of the dialysate and replacement fluids;
	\item the  dialyzing filter, which is the site were mass and volume exchange take place.
\end{enumerate}
In the ``patient'' system, the mass exchange is two-compartmental (the intracellular compartment and the extracellular one, the latter formed by the plasma compartment plus the interstitial one). The exchange volume is three-compartmental (plasma, interstitial cells).

The model, implemented in the software for the numerical calculation Matlab\textsuperscript\textregistered{} 7 (R14), requires the input of initial data from the patient and from the dialysis machine.

Three parameters were introduced and analyzed: the mass transfer coefficient to the cell membrane $k$, the dimensionless parameter $\rho$, accounting for the variability  of permeability in capillary walls, and the coefficient $\eta$, defined as the dialyzer extraction capability, which is dependent on the thickness of protein layer formed along the filter capillaries throuout the dialysis. Sensitivity analysis was also performed in order to understand which one of this parameters affect most the output of the model. In order to assess sensitivity, parameters' values were individually changed ($\pm 10\%$ of their reference values) while keeping constant all the others and variations in the output were recorded. Outcoming results show that the parameter for a given solute influences both, predictably, the dynamics of the solute itself, but also that of other solutes, thus highlighting the fact that the model has been developed with solutes coexisting and interacting in the same environment.

One of the ultimate goals was to evaluate  for how long the parameters identified in the first session would be valid for the prediction of the following ones, so that we can provide a useful tool for medical simulation. The model presented here shows overall results in line and close to those obtained by clinical observations and can therefore be considered, with few exceptions, a valid tool for prediction. In fact, deviations from real plasma concentrations of almost all solutes considered here, are at most $20\%$. In particular, the best results were obtained for sodium, potassium, chlorine, calcium and magnesium, with deviation of at most $10\%$. If we consider the average values, all the descriptive errors falls below $10\%$. Detailed analysis for every solute is given in the last chapters, along with some comments on how to improve the model descriptive capabilities.

Eventually, it is shown how the model can be used to evaluate the differences between the available dilution techniques, and their effect on the dynamics of solutes during dialysis. It was found that the degree of compartmentalization and the electrical charge of solutes may determine which of the two modes of diluition (pre- or post-diluition) would be the most effective.